Predicting computational reproducibility of scientific pipelines using collaborative filtering

### 2018-07-20 paper.tex
Worked on Section II and III.A

@@

Collaborative filtering is a technique to predict unknown values of a 
matrix called ``utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented by columns, by 
users, represented by rows. Ratings might be explicit, when users input 
ratings through a dedicated system, or implicit, when users' behaviors 
are analyzed to estimate their preference toward specific items. A 
complete introduction to collaborative filtering is provided 
in~\cite{mmdsbook}.

Several methods have been proposed to implement collaborative 
filtering. Item-item collaborative filtering~\ref{item-item} predicts 
the rating of item $i$ by user $u$ from the ratings of items similar to 
$i$ by user $u$. Likewise, user-user collaborative 
filtering~\ref{user-user} predicts the rating of item $i$ by user $u$ 
from the ratings of item $i$ by users similar to $u$. Both methods
have been used extensively for e-commerce applications.

A third class of methods, which we will apply to our problem, is based 
on the factorization of the utility matrix to estimate latent factors 
with which users and items are represented. This method is described 
in~\cite{koren} and became famous as it allowed its authors to win the 
Netflix prize in 2009. To summarize, the method aims at finding $q_i$ 
and $p_u$ vectors of $\mathbb{R}^f$, where $f$ is the number of latent factors, such that:
\begin{equation*}
r_{ui} = q_i^Tp_u,
\end{equation*}
where $r_{ui}$ is the rating of item $i$ by user $u$. In practice, the optimization
involves a regularization term to avoid overfiting particular users or items. The method
finds $q_i$ and $p_u$ that minimize the following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2\right)
,
\end{equation*}
where $\lambda$ is the regularization parameter and $\mathbb{T}$ is the training set.

@@

% Explain that you have binary classes (rounding)

% Biases

@@

We investigated the following sampling techniques for the training set. 
In each method, we included the first row of the matrix (first 
generated file of each subject) and a random column (all files of a 
random subject) to avoid cold start issues.

% 0. random unreal (baseline)
\paragraph{Random Unreal}
The training set is sampled randomly with no restriction, that is,
regardless of the file generation time. This method will be used as a
baseline for comparison with other methods, although it is not
realistic. Figure~\ref{fig:Random-Unreal-Sample-Training-set} represents 
a training set sampled using this method.

% 2. columns
\paragraph{Complete Columns}
The training set is sampled by randomly selecting complete columns in
the matrix, that is, complete subject executions. The last selected
column might be incomplete to meet the exact training ratio. This 
method is realistic: it corresponds to a situation where the 
collaborative filtering method will predict the reproducibility in the 
remaining subjects from the subjects in the training set. 
Figure~\ref{fig:Columns-Sample-Training-set} represents a training set 
sampled using this method.

% 3. rows
\paragraph{Complete Rows} The training set is sampled by selecting complete 
rows in the matrix, that is, the first files produced by
every execution. The last selected row might be incomplete to meet the
exact training ratio. This method is realistic: it corresponds to a 
situation where the processing of all the subjects is launched and 
interrupted before the execution is complete. The collaborative 
filtering method will then predict the reproducibility of the remaining 
files. Figure~\ref{fig:Rows-Sample-Training-set} represents a training 
set sampled using this method.
\begin{figure}[h!]
	\centering
	\begin{subfigure}[b]{0.4\linewidth}
		\includegraphics[width=\columnwidth]{figures/5vs7_random-unreal_04_training}
  		\caption{Random Unreal method}
  		\label{fig:Random-Unreal-Sample-Training-set}
	\end{subfigure}
	\begin{subfigure}[b]{0.4\linewidth}
  		\includegraphics[width=\columnwidth]{figures/5vs7_columns_04_training}
  		\caption{Complete Columns method}
  		\label{fig:Columns-Sample-Training-set}
	\end{subfigure}
	\begin{subfigure}[b]{0.4\linewidth}
  		\includegraphics[width=\columnwidth]{figures/5vs7_rows_04_training}
  		\caption{Complete Rows method}
  		\label{fig:Rows-Sample-Training-set}
	\end{subfigure}
	\begin{subfigure}[b]{0.4\linewidth}
 		\includegraphics[width=\columnwidth]{figures/5vs7_random-real_04_training}
		\caption{RS method}
  		\label{fig:Uniform-S-Sample-Training-set}
	\end{subfigure}
	\begin{subfigure}[b]{0.4\linewidth}
  		\includegraphics[width=\columnwidth]{figures/5vs7_diagonal_04_training}
  		\caption{RFNU  method }
  		\label{fig:Diagonal-Sample-Training-set}
	\end{subfigure}
	\begin{subfigure}[b]{0.4\linewidth}
  		\includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-largest_04_training}
  		\caption{RFNT-L method}
  		\label{fig:triangular-L-Sample-Training-set}
	\end{subfigure}
	\begin{subfigure}[b]{0.4\linewidth}
  		\includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-smallest_04_training}
  		\caption{RFNT-S method}
  		\label{fig:triangular-S-Sample-Training-set}
	\end{subfigure}
	\caption{Sample training set for different methods with a training ratio of 0.4.}		
\end{figure}

\paragraph{Random Subjects -- RS} This method builds the 
training set by selecting the files from random subjects 
until the training ratio is reached. The file selected in a 
subject is the file with the lowest index in this subject that has not 
been already selected in the training set. This method is realistic as 
files are sampled according to their production time-stamps. 
Figure~\ref{fig:Uniform-S-Sample-Training-set} represents a training 
set sampled using this method.

\paragraph{Random File Numbers (Uniform) -- RFNU}

The number of files selected for every subject is randomly selected in
a uniform distribution $U(\textit{a},\textit{b})$, where \textit{b} is set to the total
number of files $N_{f}$ and \textit{a} is set according to training ratio $\alpha$ as follows:

\[
  \begin{cases}
          \textit{a} = 0      & \text{if $\alpha \leq 0.5$ }\\

          \textit{a} = (2\alpha - 1) N_{f} & \text{if $\alpha > 0.5$}
  \end{cases}
\]

This method is realistic as files are sampled according to their production time-stamps.
Due to sampling issues, it is possible that the actual training ratio obtained with this method
does not match the target one. We check that the difference between the target and real training ratios
was lower than 0.01.\\

\paragraph{Random File Numbers (Triangular) -- RFNT}
The number of files selected for every subject is randomly selected in
a triangular distribution with parameters \textit{a}, \textit{b} and 
\textit{c} (\todo{see Figure X}). The mean of the distribution is 
$\frac{a+b+c}{3}$. In our case, we set \textit{c} to $N_{f}$ and we set 
\textit{a} and \textit{b} with two approaches meant to ensure that the 
average of the distribution is $\alpha N_{f}$ ($\alpha$ is the training 
ratio):
\begin{enumerate}
        \item Largest a (RFNT-L): a is set to the 
        largest possible value, i.e., b. The average of the 
        distribution is $\frac{2a+N_{f}}{3}$, therefore 
        $a=\frac{3\alpha-1}{2}N_{f}$.
                \item Smallest a (RFNT-S): a is 
        set to the smallest possible value, i.e., 0. The average of the 
        distribution is $\frac{b+N_{f}}{3}$, therefore 
        $b=N_{f}(3\alpha-1)$. To ensure that $b<N_{f}$, we actually set 
        $b=\min(N_{f}, N_{f}(3\alpha-1))$. 
\end{enumerate}
It should be noted that $\alpha$ must be larger than $\frac{1}{3}$ for 
a and b to remain positive. For training ratios lower than $\frac{1}{3}$, we 
define RFNT methods as the RFNU one presented previously.

As illustrated in Figures~\ref{fig:triangular-L-Sample-Training-set} 
and~\ref{fig:triangular-S-Sample-Training-set}, the motivation for the 
RFNT-L method is to guarantee that, for large enough values of 
$\alpha$, all subjects will have at least a few files in the training 
set.

@@

>>>>

\todo{Explain that columns represent subjects and rows represent files, and that these terms are used
indifferently}

@@

Collaborative filtering is a technique to predict unknown values of a 
matrix called ``utility matrix" from the known ones. Traditionally, the 
matrix represents the ratings of items, represented in columns, by 
users, represented in rows. Ratings might be explicit, when users 
provide ratings through a dedicated system, or implicit, when users' 
behaviors are tracked and analyzed to estimate their preferences. An 
overview of collaborative filtering is provided 
in~\cite{leskovec2014mining}.

Several methods have been proposed to implement collaborative 
filtering. Item-item collaborative filtering~\cite{breese1998empirical, linden2003amazon} predicts 
the rating of item $i$ by user $u$ from the ratings of items similar to 
$i$ by user $u$. Likewise, user-user collaborative 
filtering~\cite{breese1998empirical} predicts the rating of item $i$ by user $u$ 
from the ratings of item $i$ by users similar to $u$. Both methods
have been used extensively for e-commerce applications.

A third class of methods, which is the one we will apply to our 
problem, is based on the factorization of the utility matrix to 
estimate latent factors along which users and items are represented. 
This method is described 
in~\cite{koren2009matrix} and became famous as it contributed to winning the 
Netflix prize in 2009. To summarize, the method aims at finding $q_i$ 
and $p_u$ vectors of $\mathbb{R}^f$, where $f$ is the number of latent factors, such that:
\begin{equation*}
r_{ui} = q_i^Tp_u,
\end{equation*}
where $r_{ui}$ is the rating of item $i$ by user $u$. In practice, the optimization
involves a regularization term to avoid overfiting particular users or items. The method
finds $q_i$ and $p_u$ that minimize the following objective:
\begin{equation*}
\sum_{(i,u) \in \mathbb{T}}\left( r_{ui} - q_i^Tp_u\right)^2+\lambda \left( \|{q_i}\|^2 + \|{p_u}\|^2\right)
,
\end{equation*}
where $\lambda$ is the regularization parameter and $\mathbb{T}$ is the training set.

@@

% Explain our particular context.
% What our matrix represents
% Explain that you have binary classes (rounding)
% Explain the sampling issue.

@@

As explained before, the training set in our problem cannot be 
constructed by unconstrained random sampling of the matrix.
To address this issue, we investigated the following sampling techniques.

% 2. columns
\paragraph{Complete Columns (Figure~\ref{fig:Columns-Sample-Training-set})}
The training set is sampled by randomly selecting complete columns in
the matrix, that is, complete subject executions. The last selected
column might be incomplete to meet the exact training ratio. This 
method is realistic: it corresponds to a situation where the 
collaborative filtering method will predict the reproducibility in the 
remaining subjects from the subjects in the training set.

% 3. rows
\paragraph{Complete Rows (Figure~\ref{fig:Rows-Sample-Training-set})} The training set is sampled by selecting complete 
rows in the matrix, that is, the first files produced by
every execution. The last selected row might be incomplete to meet the
exact training ratio. This method is realistic: it corresponds to a 
situation where the processing of all the subjects is launched and 
interrupted before the execution is complete. The collaborative 
filtering method will then predict the reproducibility of the remaining 
files.

\begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_columns_04_training}
                  \caption{Complete Columns method}
                  \label{fig:Columns-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_rows_04_training}
                  \caption{Complete Rows method}
                  \label{fig:Rows-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                 \includegraphics[width=\columnwidth]{figures/5vs7_random-real_04_training}
                \caption{RS method}
                  \label{fig:Uniform-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_diagonal_04_training}
                  \caption{RFNU  method }
                  \label{fig:Diagonal-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-largest_04_training}
                  \caption{RFNT-L method}
                  \label{fig:triangular-L-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{figures/5vs7_random-triangular-smallest_04_training}
                  \caption{RFNT-S method}
                  \label{fig:triangular-S-Sample-Training-set}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
        \includegraphics[width=\columnwidth]{figures/5vs7_random-unreal_04_training}
                  \caption{Random Unreal method}
                  \label{fig:Random-Unreal-Sample-Training-set}
        \end{subfigure}
        \caption{Sample training set for different methods with a training ratio of 0.4.}                
\end{figure}

\paragraph{Random Subjects -- RS (Figure~\ref{fig:Uniform-S-Sample-Training-set})} This method builds the 
training set by selecting the files from random subjects 
until the training ratio is reached. The file selected in a 
subject is the file with the lowest index in this subject that has not 
been already selected in the training set. This method is realistic as 
files are sampled according to their production time-stamps.

\paragraph{Random File Numbers (Uniform) -- RFNU (Figure~\ref{fig:Diagonal-Sample-Training-set})}
The number of files selected for every subject is randomly selected in
a uniform distribution $U(\textit{a},\textit{b})$, where \textit{b} is set to the total
number of files $N_{f}$ and \textit{a} is set according to training ratio $\alpha$ as follows:
\[
  \begin{cases}
          \textit{a} = 0      & \text{if $\alpha \leq 0.5$ }\\

          \textit{a} = (2\alpha - 1) N_{f} & \text{if $\alpha > 0.5$}
  \end{cases}
\]
This method is realistic as files are sampled according to their production time-stamps.
Due to sampling issues, it is possible that the actual training ratio obtained with this method
does not match the target one. We check that the difference between the target and real training ratios
was lower than 0.01.\\

\paragraph{Random File Numbers (Triangular) -- RFNT}
The number of files selected for every subject is randomly selected in
a triangular distribution with parameters \textit{a}, \textit{b} and 
\textit{c} as in Figure~\ref{fig:triangular}. The mean of the distribution is 
$\frac{a+b+c}{3}$. We set \textit{c} to $N_{f}$ and we set 
\textit{a} and \textit{b} with two approaches that ensure that the 
average of the distribution is $\alpha N_{f}$ ($\alpha$ is the training 
ratio):
\begin{enumerate}
        \item Largest a (RFNT-L, Figure~\ref{fig:triangular-L-Sample-Training-set}): a is set to the 
        largest possible value, i.e., b. The average of the 
        distribution is $\frac{2a+N_{f}}{3}$, therefore 
        $a=\frac{3\alpha-1}{2}N_{f}$.
                \item Smallest a (RFNT-S, Figure~\ref{fig:triangular-S-Sample-Training-set}): a is 
        set to the smallest possible value, i.e., 0. The average of the 
        distribution is $\frac{b+N_{f}}{3}$, therefore 
        $b=N_{f}(3\alpha-1)$. To ensure that $b<N_{f}$, we actually set 
        $b$ to $\min(N_{f}, N_{f}(3\alpha-1))$. 
\end{enumerate}
It should be noted that $\alpha$ must be larger than $\frac{1}{3}$ for 
a and b to remain positive. For training ratios lower than $\frac{1}{3}$, we 
define RFNT methods as the RFNU one presented previously.
As illustrated in Figures~\ref{fig:triangular-L-Sample-Training-set} 
and~\ref{fig:triangular-S-Sample-Training-set}, the motivation for the 
RFNT-L method is to guarantee that, for large enough values of 
$\alpha$, all subjects will have at least a few files in the training 
set, which is not the case for RFNU.
\begin{figure}
\centering
\includegraphics[width=0.5\columnwidth]{figures/triangular.pdf}
\caption{Triangular distribution}
\label{fig:triangular}
\end{figure}

% 0. random unreal (baseline)
\paragraph{Random Unreal (Figure ~\ref{fig:Random-Unreal-Sample-Training-set})}
The training set is sampled in a random uniform way, 
regardless of the file generation times. This method will be used as a
baseline for comparison with other methods, although it is not
realistic.

In each method, we also included the first row of the matrix (first 
generated file of each subject) and a random column (all files of a 
random subject) to avoid cold start issues.

@@

\subsection{Synthetic Data}

We generated synthetic matrices as shown in 
Figure~\ref{fig:synthetic-data}. Each matrix has 100 files and 100 
subjects of different \emph{types}. Subjects of the same type behave 
identically and all types contain the same number of subjects $\pm$ 1 
\todo{Soudabeh: check that}. Such a decomposition by subject type 
correspond to the case where different subjects may have data of 
different nature, as it is for instance the case in the neuroimaging 
dataset of the Human Connectome Project~\cite{hcp} where not all the 
subjects have the same amount of images.

For subjects of a given type, matrix elements vary in $log(n)$  blocks, 
where $n$ is the number of types, with all the possible variation 
patterns: some types do not vary at all, while other ones vary in every 
block. Such variation patterns are meant to mimic the logic of data 
processing pipelines: each block of files represents the files produced 
at a given stage of the pipeline, which may or may not contain 
reproducibility errors depending on the subject type.

\begin{figure}
\centering
        %\begin{subfigure}[b]{0.4\linewidth}
                  %\includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/1_SubjectType_utility_matrix.png}
                  %\caption{1 type}
        %\end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/2_SubjectType_utility_matrix.png}
                  \caption{2 types}
        \end{subfigure}
        \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/4_SubjectType_utility_matrix.png}
                  \caption{4 types}
        \end{subfigure}
                \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/8_SubjectType_utility_matrix.png}
                  \caption{8 types}
        \end{subfigure}
                \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/16_SubjectType_utility_matrix.png}
                  \caption{16 types}
        \end{subfigure}
                \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/32_SubjectType_utility_matrix.png}
                  \caption{32 types}
        \end{subfigure}
                \begin{subfigure}[b]{0.4\linewidth}
                  \includegraphics[width=\columnwidth]{data/Utility_Matrix/Synthetic/synthetic_subject_types/64_SubjectType_utility_matrix.png}
                  \caption{64 types}
        \end{subfigure}
\caption{Synthetic Data}
\label{fig:synthetic-data}
\end{figure}

\subsection{Real Data}
